# Chapter 5: Visualisation: good cartographic design

## Introduction

This chapter aims to focus on introducing good practice in map design and presentation. When putting a map together you need to think about its intended audience (their level of expertise, whether you want them to interact with the map), purpose, and format of delivery (e.g., printed, web, projected in a screen, etc). There are many design decisions you need to consider: fonts, labels, colour, legends, layout, etc. In this chapter we provide a general introduction to some basic design principles for map production. These themes, and the appropriate election of symbol representation, are the subject matter of cartography, the art and science of map making. Within cartography a considerable body of research and scholarship has focused on studying the visual and psychological implications of our mapping choices. As noted on previous chapters one of the problems with maps is that powerful as a tool as they, they can lead to misunderstanding. What the mapmaker chooses to emphasise and what the map reader see may not be the same thing. We will work you through an example of a fairly basic map and the process of taking to a point where it could be ready for presentation to an audience other that yourself.


In this chapter we will be working with some data published by Hungarian police available online [http://www.police.hu/hu/a-rendorsegrol/statisztikak/kozrendvedelem](http://www.police.hu/hu/a-rendorsegrol/statisztikak/kozrendvedelem). Specifically we will be looking at some statistics related to drink driving. Drunk driving is one of a number of problems police confront that relate to impaired and dangerous driving. Hungary has a strict drink driving policy, with the maximum drink diving limit being 0.0 BAC. Most European countries are at 0.5 BAC, while the UK is 0.8 (except 0.5 for Scotland). We have record for each county the number of brethalyser checks carried out, and the number of these which returned a positive result. Let's read in this data from January 2020 (taken from [here](http://www.police.hu/sites/default/files/Kozrendvedelem%20SK%202021_01.pdf)), 


```{r}

library(sf)
library(readr)
library(dplyr)



hungary <- st_read("data/hungary.geojson")
drink_driving <- read_csv("data/drink_driving.csv")

hu_dd <- left_join(hungary, drink_driving, by = c("name" = "name"))

```


We can now use this example to talk through the important principles of good visualisation of spatial data. We draw specifically from two areas of research: cartography and data visualisation. 

Cartographers have always been concerned about the appearance of maps and how the display marries form with function @Field_2012. As there is no definitive definition for what is meant by *cartographic design* it can be challenging to evaluate what makes *good* design. However there are themes and elements which can be used to guide the map maker, and offer points of reflection to encourage thoughtful designs. 

Te primary aim of maps is the communication of information in an honest and ethical way. This means each map should have a clear goal and know its audience, show all relevant data and not use the data to lie or mislead @Dent_2009. It should also be reproducible, transparent, cite all data sources, and consider diversity in its audience @Dent_2009. So what does that mean for specifically implementing these into practice. While a good amount of critical thought from the map maker will be required, there are aids we can rely upon. For example, @Field_2007 developed a map evaluation checklist which you can access here: [http://downloads.esri.com/MappingCenter2007/arcGISResources/more/MapEvaluationGuidelines.pdf](http://downloads.esri.com/MappingCenter2007/arcGISResources/more/MapEvaluationGuidelines.pdf). The questions fall into three broad categories: *Cartographic Requirements* such as what is the rationale for the map, who are the audience?. *Cartographic Complication and Design* such as are all the relevant features included and do the colours, symbold, and other features legible and appropriate to achieve the map's objectives? And finally, *Map Elements and Page Layout* which tackle some specific features such as orientation indicator, scale indicator, legend, titles and subtitles, and production notes. We will discuss these elements in this chapter to some degree, and the recommended reading will guide the reader to further advice on these topics.    

Data visualisation is a somewhat newer field, however, it seems to encompass the same guiding principles when consdiering what makes good design.  According to @Kirk_2016 three principles offer a guide when deciding what makes a *good* data visualisation. It must be: 

- Trustworthy
- Accessible, and 
- Elegant

The first principle, of trust speaks to the integrity, accuracy, and legitimacy of any data visualisation we produce. @Kirk_2016 suggests this principle to be held above all else, as our primary goal is to communicate truth (as far as we know it) and avoid at all cost to present what we know to be misleading content.  Accessibility refers to our visualisation being useful, understandable, and unobtrusive, as well as accessible for all users. There are many things to consider in your audience such as *dynamic of need* (do they have to engage with your visualisation, or is it voluntary?), *subject-matter knowledge* are they experts in the area, are they lay people to whom you must communicate a complex message?, and many other factors (see @Kirk_2016). Finally, elegance refers to aesthetics, attention to detail, and an element of *doing as little design as possible* - meaning a certain invisibility whereby the viewer of your visualisation focussed on the content, rather than the design - that is the main point is the message that you are trying to communicate with your data!


In this chapter, we will aim to bring together the above principles, and work through a practical example of how to apply this to achieve good visualisations. Specifically we will cover: 

- data representation
- colour
- text
  + titles and subtitles
  + legend
  + annotation
  + production notes
- composition
   + orientation indicator
   + scale indicator
   + borders
- inset maps 
- interactivity (??)

## Data representation


### Thematic maps

We've been working with thematic maps thus far. There are many decisions that go into making a thematic map, which we have explored at length in the previous chapters, such as how (and whether) to bin your data (Chapter 3) and how (or whether) to transform your polygons (Chapter 4). These are important considerations on how to represent your data to your audience, and require a technical understanding, not only an aesthetic one. So please do read over those chapters carefully when thinking about how to represent your data. 


We can use the `ggplot2` package to plot our thematic map, using the `geom_sf()` function. To shade each polygon with the values of a specific variable, we use the `fill = ` argument within the `aes()` (aesthetics) function. Most simply: 

```{r themmap}

library(ggplot2)

ggplot(data = hu_dd) + 
  geom_sf(aes(fill = total_breath_tests)) + 
  theme_minimal()

```

We can change the colour and size of the borders of our polygons with arguments inside the `geom_sf()` function, but outside the `aes()` function, as long as we're not using our data to define these. For example we can change the line width (`lwd =`) to 0, eliminating bordering lines between our polygons: 


```{r themmapnobord}

ggplot(data = hu_dd) + 
  geom_sf(aes(fill = total_breath_tests), lwd = 0) + 
  theme_minimal()

```


Or we can change the colour of the borders with the `col =` argument: 



```{r themmapwhitebord}

ggplot(data = hu_dd) + 
  geom_sf(aes(fill = total_breath_tests), lwd = 0.5, col = "white") + 
  theme_minimal()

```

Here we have a continuous fill for our values, however we can employ our learning from Chapter 3 and apply a classification system, such as quantiles. To do this we might create a new variable which contains the quantiles of our numeric variable, and then use that as our `fill = `. 

```{r}

hu_dd <- hu_dd %>% 
  mutate(total_quantiles = cut(total_breath_tests, breaks = round(quantile(total_breath_tests),0), include.lowest = TRUE, dig.lab=10))


ggplot(data = hu_dd) + 
  geom_sf(aes(fill = total_quantiles), lwd = 0.5, col = "white") + 
  theme_minimal()

```

The colour scheme is terrible, but we will talk about colour in the next section, so we can forgive that for now...

### Symbols

You might not want to display your map as thematic map, you may want to use symbols. Again we explored this in Chapter 3, here is one way you can use graduated symbol map with `ggplot()`. You can take the centroid of each county polygon using the `st_centroid()` function from the `sf` package, and then when  mapping with `geom_sf()`, within the `aes()` function specity the `size = ` argument to the variable you wish to visualise: 

```{r gradsymbmap}

ggplot(data = hu_dd) + 
  geom_sf() + 
  geom_sf(data = st_centroid(hu_dd), aes(size = total_breath_tests)) + 
  theme_minimal()

```
Like with the thematic map you can play around with colour and shape: 

```{r gradsymbmapcol}

ggplot(data = hu_dd) + 
  geom_sf(fill = "light yellow", col = "white") + 
  geom_sf(data = st_centroid(hu_dd), aes(size = total_breath_tests), col = "orange") + 
  theme_minimal()

```

```{r gradsymbmapcoltriangle}

ggplot(data = hu_dd) + 
  geom_sf(fill = "light yellow", col = "white") + 
  geom_sf(data = st_centroid(hu_dd), aes(size = total_breath_tests), col = "orange", shape = 17) + 
  theme_minimal()

```

```{r gradsymbmapcoldiamond}

ggplot(data = hu_dd) + 
  geom_sf(fill = "light yellow", col = "white") + 
  geom_sf(data = st_centroid(hu_dd), aes(size = total_breath_tests), col = "orange", shape = 13) + 
  theme_minimal()

```

How you choose to represent your data will depend on your decisions to the questions asked above about audience, message, integrity, and so on. 

### Rate v count

In chapter 3 we have discussed this already in great detail, so again I will not dwell on this, but it is important that your data are meaningful and easy to interpret. We might, in this case for example, want to consider the rate of positive breath tests per test carried out in each county. To compute this, we might want to consider the proportion of positive results on the breatalyser tests (where the person had been drinking and their result is over the limit). To compute this, we can simply divide the positive results by the total test, and multiply by 100. We also include the `round()` function in there 


```{r}

hu_dd <- hu_dd %>% 
  mutate(pos_rate = round(positive_breath_tests/total_breath_tests*100,1))


```


We can see the county with the highers proportion of test yielding drink drivers is `r hu_dd %>% filter(pos_rate == max(pos_rate)) %>% pull(name)` with `r hu_dd %>% filter(pos_rate == max(pos_rate)) %>% pull(pos_rate)` %, while the county with the lowest is `r hu_dd %>% filter(pos_rate == min(pos_rate)) %>% pull(name)` with `r hu_dd %>% filter(pos_rate == min(pos_rate)) %>% pull(pos_rate)` %. 


We can visualise this rate on our thematic map in exactly the same way as the count data, but using our new variable in the `fill =` argument:  

```{r themmapwhitebord}

ggplot(data = hu_dd) + 
  geom_sf(aes(fill = pos_rate), lwd = 0.5, col = "white") + 
  theme_minimal()

```


But with a graduated symbol map we can get a little more creative. Sure, one approach may be to once again change the size of the symbol, but we could get a little more creative and use charts. For example there is the library `scatterpie` which has the function `geom_scatterpie()` which allows us to present our ratios in pie charts. 


To use this function, we need our data to be in a dataframe which has separate columns for latitude and longitude (rather than a geometry object), where each observation has a unique id, and where we have a separate column for each proportion we wish to visualise. 


To create this, we first need to get the centroid of each polygon to get our longitude and latitude. We also need to calculate the number of negative breath tests. We then need to extract the coordinates into longitude and latitude columns, and finally we can drop the geometry and keep only the columns we need for this new dataframe. 



```{r}


scatter_map_df <- hu_dd %>% 
  mutate(cent_lng = st_coordinates(st_centroid(.))[,1],    # extract centroid longitude
         cent_lat = st_coordinates(st_centroid(.))[,2],    # extract centroid latitude
         neg_tests = total_breath_tests - positive_breath_tests) %>%     # calculate negative tests
  st_drop_geometry() %>%       # remove the geometry column
  select(cent_lng, cent_lat, osm_id, positive_breath_tests, neg_tests)   # keep only necessary columns

```


We now have this new dataframe which has coordinates for the centroid of each polygon, and we can use the `geom_scatterpie()` function from the `scatterpie` package to plot a pie chart of the proportion of positive to negative breath tests in each county in Hungary: 


```{r}
ggplot(data = hu_dd) + 
  geom_sf(data = hu_dd, fill = "light yellow", col = "white") + 
  scatterpie::geom_scatterpie(data=scatter_map_df, aes(x=cent_lng, y=cent_lat, group=osm_id), cols=c("positive_breath_tests", "neg_tests")) + 
  theme_minimal()

```

In this case, it's not super informative to visualise this, as discussed above our positive rates range between  `r hu_dd %>% filter(pos_rate == min(pos_rate)) %>% pull(pos_rate)` %, and `r hu_dd %>% filter(pos_rate == max(pos_rate)) %>% pull(pos_rate)` %. However with more evenly distributed proportions this may be something to try. 

## Colour

When choosing a colour palette, the first thing to consider is what kind of colour scheme we need. This will depend on the variable we are trying to visualise. We go back, once again, to the first week of the course, where we discussed *Levels of Measurement*. Remember those? Still important!


Depending on the kind of varaible we want to visualise, we might want a Qualitative colour scheme (for categorical nominal variables), a Sequential colour scheme (for categorial ordinal, or for numeric variables) or a Diverging colour scheme (for categorial ordinal, or for numeric variables). 

For qualitative colour schemes, we want each category (each value for the variable) to have a perceptible difference in colour. For sequential and diverging color schemes, we will want mappings from data to color that are not just numerically but also perceptually uniform. 



- **sequential scales** (also called gradients) go from low to high saturation of a colour. 
- **diverging scales** represent a scale with a neutral mid-point (as when we are showing temperatures, for instance, or variance in either direction from a zero point or a mean value), where the steps away from the midpoint are perceptually even in both directions. 
- **qualitative scales** identify as different the different values of your categorical nominal variable from each other. 



For your sequetial and diverging scales, the goal in each case is to generate a perceptually uniform scheme, where hops from one level to the next are seen as having the same magnitude. 

Of course, perceptual uniformity matters for your qualitative scales for your unordered categorical variables as well. We often use color to represent data for different countries, or political parties, or types of people, and so on. In those cases we want the colors in our qualitative palette to be easily distinguishable, but also have the same valence for the viewer. Unless we are doing it deliberately, we do not want one color to perceptually dominate the others. 



The main message here is that you should generally not put together your color palettes in an ad hoc way. It is too easy to go astray. In addition to the considerations we have been discussing, there we might also want to avoid producing plots that confuse people who are colour blind, for example, and color blindness comes in a variety of forms. Fortunately for us, almost all of the work has been done for us already. Different color spaces have been defined and standardized in ways that account for these uneven or nonlinear aspects of human color perception.


A good resource is [colorbrewer](http://colorbrewer2.org/). We have come across @Brewer_2016 in Chapter 3. Colorbrewer is a resource developed by Cynthia Brewer and colleagues in order to help implement good colour practice in data visualisation and cartography.  This site offers many colour schemes we can make use of for our maps, which are easily integrated into R using the `Rcolorbrewer` package.


```{r, loadrcolorbrewer}
library(RColorBrewer)

```


Once you have the package loaded, we can look at all the associated palettes with the function `display.brewer.all()`.

```{r, dsplayallcolourbrewerpallettes}

display.brewer.all()

```

The above gives a wide choice of pallettes, and while they are applicable to all sorts of data visualisations, they were created especially for the case of thematic maps. 


We might use the above code to pick a palette we like. We might then want to examine the colours more closely. To do this we can use the `display.brewer.pal()` function, and specify `n=` - the number of colours we need, as well as the palette name with `name = `: 


```{r}

display.brewer.pal(n = 5, "Spectral")

```

Let's go back to our thematic map of the quantiles of total breath tests per county. We might be interested in this map to show distribution of policing activity for example. We made this map earlier with the default colour scheme, which didn't really communicate to use the graduated nature of our data we were visualising. To properly do this, we may imagine using a *seqantial* scale. We can use one of the sequential scales available within `RColorBrewer` with adding the `scale_fill_brewer()` function to our ggplot. In this function we can specify the `type=` parameter, i.e. if we want to use sequential, divergining, or qualitative colour schemes (specified as either "seq" (sequential), "div" (diverging) or "qual" (qualitative)). We can then specify our preferred palette with the `palette = ` argument. 

Let's demonstrate here with the "YlOrRd" sequential palette: 

```{r, colourbbrewerratefill}

ggplot(data = hu_dd) + 
  geom_sf(aes(fill = total_quantiles), lwd = 0.5, col = "white") + 
  scale_fill_brewer(type = "seq", palette = "YlOrRd") + 
  theme_minimal()

```

This looks much better, and communicates our message much more clearly. Is this accessible to our colourblind colleagues? Earlier, when we asked to view all the palettes with the `display.brewer.all()` function, we did not specify any arguments. However, we can do so in order to filter only those palettes which are accessible for all audiences. We can include the parameter `colorblindFriendly = ` to do so: 

```{r}

display.brewer.all(colorblindFriendly = TRUE)

```

You can see there are a few palettes missing from our earlier results, when we did not specify this requirement. My recommendation is to always use one of these palettes. 


Another way to ensure that we are making accessible maps is to use greyscale (if your map is being printed, this may also save some money). To introduce a greyscale palette, you can use the function `scale_fill_grey() ` from the `ggplot2` package: 


```{r greyscaletestsmap}

ggplot(data = hu_dd) + 
  geom_sf(aes(fill = total_quantiles), lwd = 0.5, col = "white") + 
  scale_fill_grey() + 
  theme_minimal()

```


Sometimes you might prefer such a map. However, do keep in mind, a number of studies have shown the desirability of monochrome colour (over greyscale) thematic maps, as they are linked to less observer variability in interpretation (@Lawson_2006). So you might want to use something like this instead: 

```{r}

ggplot(data = hu_dd) + 
  geom_sf(aes(fill = total_quantiles), lwd = 0.5, col = "white") + 
  scale_fill_brewer(type = "seq", palette = "Greens") + 
  theme_minimal()

```

Overall, the key thing is to be conscious with the colours you choose to represent your data. Make sure that they are accessible for all audiences, and best represent the patterns in your data which you want to communicate to your audiences. 


## Text

### Titles and subtitles

### Legend

### Annotation

### production notes

## Composition

### Orientation indicator

### Scale indicator

### Borders


## Inset maps

(INSET MAPS)
https://geocompr.github.io/post/2019/ggplot2-inset-maps/

## Interactivity
