# Chapter 5: Visualisation: good cartographic design

## Introduction

This chapter aims to focus on introducing good practice in map design and presentation. When putting a map together you need to think about its intended audience (their level of expertise, whether you want them to interact with the map), purpose, and format of delivery (e.g., printed, web, projected in a screen, etc). There are many design decisions you need to consider: fonts, labels, colour, legends, layout, etc. In this chapter we provide a general introduction to some basic design principles for map production. These themes, and the appropriate election of symbol representation, are the subject matter of cartography, the art and science of map making. Within cartography a considerable body of research and scholarship has focused on studying the visual and psychological implications of our mapping choices. As noted on previous chapters one of the problems with maps is that powerful as a tool as they, they can lead to misunderstanding. What the mapmaker chooses to emphasise and what the map reader see may not be the same thing. We will work you through an example of a fairly basic map and the process of taking to a point where it could be ready for presentation to an audience other that yourself.


In this chapter we will be working with some data published by Hungarian police available online [http://www.police.hu/hu/a-rendorsegrol/statisztikak/kozrendvedelem](http://www.police.hu/hu/a-rendorsegrol/statisztikak/kozrendvedelem). Specifically we will be looking at some statistics related to drink driving. Drunk driving is one of a number of problems police confront that relate to impaired and dangerous driving. Hungary has a strict drink driving policy, with the maximum drink diving limit being 0.0 BAC. Most European countries are at 0.5 BAC, while the UK is 0.8 (except 0.5 for Scotland). We have record for each county the number of brethalyser checks carried out, and the number of these which returned a positive result. Let's read in this data from January 2020 (taken from [here](http://www.police.hu/sites/default/files/Kozrendvedelem%20SK%202021_01.pdf)), 


```{r}

library(sf)
library(readr)



hungary <- st_read("data/hungary.geojson")
drink_driving <- read_csv("data/drink_driving.csv")

hu_dd <- left_join(hungary, drink_driving, by = c("name" = "name"))

```


We can now use this example to talk through the important principles of good visualisation of spatial data. We draw specifically from two areas of research: cartography and data visualisation. 

Cartographers have always been concerned about the appearance of maps and how the display marries form with function @Field_2012. As there is no definitive definition for what is meant by *cartographic design* it can be challenging to evaluate what makes *good* design. However there are themes and elements which can be used to guide the map maker, and offer points of reflection to encourage thoughtful designs. 

Te primary aim of maps is the communication of information in an honest and ethical way. This means each map should have a clear goal and know its audience, show all relevant data and not use the data to lie or mislead @Dent_2009. It should also be reproducible, transparent, cite all data sources, and consider diversity in its audience @Dent_2009. So what does that mean for specifically implementing these into practice. While a good amount of critical thought from the map maker will be required, there are aids we can rely upon. For example, @Field_2007 developed a map evaluation checklist which you can access here: [http://downloads.esri.com/MappingCenter2007/arcGISResources/more/MapEvaluationGuidelines.pdf](http://downloads.esri.com/MappingCenter2007/arcGISResources/more/MapEvaluationGuidelines.pdf). The questions fall into three broad categories: *Cartographic Requirements* such as what is the rationale for the map, who are the audience?. *Cartographic Complication and Design* such as are all the relevant features included and do the colours, symbold, and other features legible and appropriate to achieve the map's objectives? And finally, *Map Elements and Page Layout* which tackle some specific features such as orientation indicator, scale indicator, legend, titles and subtitles, and production notes. We will discuss these elements in this chapter to some degree, and the recommended reading will guide the reader to further advice on these topics.    

Data visualisation is a somewhat newer field, however, it seems to encompass the same guiding principles when consdiering what makes good design.  According to @Kirk_2016 three principles offer a guide when deciding what makes a *good* data visualisation. It must be: 

- Trustworthy
- Accessible, and 
- Elegant

The first principle, of trust speaks to the integrity, accuracy, and legitimacy of any data visualisation we produce. @Kirk_2016 suggests this principle to be held above all else, as our primary goal is to communicate truth (as far as we know it) and avoid at all cost to present what we know to be misleading content.  Accessibility refers to our visualisation being useful, understandable, and unobtrusive, as well as accessible for all users. There are many things to consider in your audience such as *dynamic of need* (do they have to engage with your visualisation, or is it voluntary?), *subject-matter knowledge* are they experts in the area, are they lay people to whom you must communicate a complex message?, and many other factors (see @Kirk_2016). Finally, elegance refers to aesthetics, attention to detail, and an element of *doing as little design as possible* - meaning a certain invisibility whereby the viewer of your visualisation focussed on the content, rather than the design - that is the main point is the message that you are trying to communicate with your data!


In this chapter, we will aim to bring together the above principles, and work through a practical example of how to apply this to achieve good visualisations. Specifically we will cover: 

- data representation
- colour
- text
  + titles and subtitles
  + legend
  + annotation
  + production notes
- composition
   + orientation indicator
   + scale indicator
   + borders
- inset maps 
- interactivity (??)

## Data representation


### Thematic maps

We've been working with thematic maps thus far. There are many decisions that go into making a thematic map, which we have explored at length in the previous chapters, such as how (and whether) to bin your data (Chapter 3) and how (or whether) to transform your polygons (Chapter 4). These are important considerations on how to represent your data to your audience, and require a technical understanding, not only an aesthetic one. So please do read over those chapters carefully when thinking about how to represent your data. 


We can use the `ggplot2` package to plot our thematic map, using the `geom_sf()` function. To shade each polygon with the values of a specific variable, we use the `fill = ` argument within the `aes()` (aesthetics) function. Most simply: 

```{r themmap}

ggplot(data = hu_dd) + 
  geom_sf(aes(fill = total_breath_tests)) + 
  theme_minimal()

```

We can change the colour and size of the borders of our polygons with arguments inside the `geom_sf()` function, but outside the `aes()` function, as long as we're not using our data to define these. For example we can change the line width (`lwd =`) to 0, eliminating bordering lines between our polygons: 


```{r themmapnobord}

ggplot(data = hu_dd) + 
  geom_sf(aes(fill = total_breath_tests), lwd = 0) + 
  theme_minimal()

```


Or we can change the colour of the borders with the `col =` argument: 



```{r themmapwhitebord}

ggplot(data = hu_dd) + 
  geom_sf(aes(fill = total_breath_tests), lwd = 0.5, col = "white") + 
  theme_minimal()

```

Here we have a continuous fill for our values, however we can employ our learning from Chapter 3 and apply a classification system, such as quantiles. To do this we might create a new variable which contains the quantiles of our numeric variable, and then use that as our `fill = `. 

```{r}

hu_dd <- hu_dd %>% 
  mutate(total_quantiles = cut(total_breath_tests, breaks = round(quantile(total_breath_tests),0), include.lowest = TRUE, dig.lab=10))


ggplot(data = hu_dd) + 
  geom_sf(aes(fill = total_quantiles), lwd = 0.5, col = "white") + 
  theme_minimal()

```

The colour scheme is terrible, but we will talk about colour in the next section, so we can forgive that for now...

### Symbols

You might not want to display your map as thematic map, you may want to use symbols. Again we explored this in Chapter 3, here is one way you can use graduated symbol map with `ggplot()`. You can take the centroid of each county polygon using the `st_centroid()` function from the `sf` package, and then when  mapping with `geom_sf()`, within the `aes()` function specity the `size = ` argument to the variable you wish to visualise: 

```{r gradsymbmap}

ggplot(data = hu_dd) + 
  geom_sf() + 
  geom_sf(data = st_centroid(hu_dd), aes(size = total_breath_tests)) + 
  theme_minimal()

```
Like with the thematic map you can play around with colour and shape: 

```{r gradsymbmapcol}

ggplot(data = hu_dd) + 
  geom_sf(fill = "light yellow", col = "white") + 
  geom_sf(data = st_centroid(hu_dd), aes(size = total_breath_tests), col = "orange") + 
  theme_minimal()

```

```{r gradsymbmapcoltriangle}

ggplot(data = hu_dd) + 
  geom_sf(fill = "light yellow", col = "white") + 
  geom_sf(data = st_centroid(hu_dd), aes(size = total_breath_tests), col = "orange", shape = 17) + 
  theme_minimal()

```

```{r gradsymbmapcoldiamond}

ggplot(data = hu_dd) + 
  geom_sf(fill = "light yellow", col = "white") + 
  geom_sf(data = st_centroid(hu_dd), aes(size = total_breath_tests), col = "orange", shape = 13) + 
  theme_minimal()

```

How you choose to represent your data will depend on your decisions to the questions asked above about audience, message, integrity, and so on. 

### Rate v count

In chapter 3 we have discussed this already in great detail, so again I will not dwell on this, but it is important that your data are meaningful and easy to interpret. We might, in this case for example, want to consider the rate of positive breath tests per test carried out in each county. To compute this, we might want to consider the proportion of positive results on the breatalyser tests (where the person had been drinking and their result is over the limit). To compute this, we can simply divide the positive results by the total test, and multiply by 100. We also include the `round()` function in there 


```{r}

hu_dd <- hu_dd %>% 
  mutate(pos_rate = round(positive_breath_tests/total_breath_tests*100,1))


```


We can see the county with the highers proportion of test yielding drink drivers is `r hu_dd %>% filter(pos_rate == max(pos_rate)) %>% pull(name)` with `r hu_dd %>% filter(pos_rate == max(pos_rate)) %>% pull(pos_rate)` %, while the county with the lowest is `r hu_dd %>% filter(pos_rate == min(pos_rate)) %>% pull(name)` with `r hu_dd %>% filter(pos_rate == min(pos_rate)) %>% pull(pos_rate)` %. 


We can visualise this rate on our thematic map in exactly the same way as the count data, but using our new variable in the `fill =` argument:  

```{r themmapwhitebord}

ggplot(data = hu_dd) + 
  geom_sf(aes(fill = pos_rate), lwd = 0.5, col = "white") + 
  theme_minimal()

```


But with a graduated symbol map we can get a little more creative. Sure, one approach may be to once again change the size of the symbol, but we could get a little more creative and use charts. 


```{r}



scatter_map_df <- hu_dd %>% 
  mutate(centroids = st_centroid(.),
         cent_coords = st_coordinates(centroids), 
         neg_tests = total_breath_tests - positive_breath_tests)



```

## Colour

A number of studies have shown the desirability of monochrome colour (over greyscale) thematic maps, as they are linked to less observer variability in interpretation (@Lawson_2006)

## Text

### Titles and subtitles

### Legend

### Annotation

### production notes

## Composition

### Orientation indicator

### Scale indicator

### Borders


## Inset maps

(INSET MAPS)
https://geocompr.github.io/post/2019/ggplot2-inset-maps/

## Interactivity
